---
title: "Fig. 4: Drought effects on predicted bacterial functional potential"
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script is used to analyse the bacterial functional potential derived from PICRUSt2 using the `ggpicrust2` package.

### 1 \| Pre-processing: Load PICRUSt2 data 

We are using the `pathway_pca()` and `pathway_gsea()` functions for the upcoming analysis. First, we need to load and subset the data based on forest type (*terra firme*, white-sand).

**Important!** PICRUSt2 files were generated within the QIIME2 environment using the `picrust2.py` pipeline. The script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.**

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readr)
library(tibble)
library(tidyverse)
library(ggprism)
library(patchwork)
library(dplyr)
library(tidyr)
library(dplyr)

## Import data 
KO_abundance = read.delim("KO_per_sample.tsv") %>% select(-"DSTB")

metadata = read_delim("BAC_silva138.2.txt", delim = "\t", escape_double = FALSE, trim_ws = TRUE)
metadata = metadata[,-1] %>% rename(sample_name = smp_id) %>% filter(sample_name != "DSTB")

## Make grouping variable
metadata$Group <- paste(metadata$site, metadata$season, sep = "_")
metadata$Group <- factor(metadata$Group, levels = c("TF_dry", "TF_wet", "WSE_dry", "WSE_wet"))

## Define subset filters
tf_samples  <- metadata$sample_name[metadata$site == "TF"]
wse_samples <- metadata$sample_name[metadata$site == "WSE"]

## Subset Picrust2::KO data
KO_tf <- KO_abundance[, c("function.", tf_samples), drop = FALSE]
KO_wse <- KO_abundance[, c("function.", wse_samples), drop = FALSE]

KO_tf = KO_tf %>% as_tibble()
KO_wse = KO_wse %>% as_tibble()

## Subset metadata for PCA
metadata_tf  <- metadata[metadata$site == "TF", , drop = FALSE]
metadata_wse <- metadata[metadata$site == "WSE", , drop = FALSE]

## Subset metadata for GSEA 
metadata_tf2  <- metadata_tf %>% column_to_rownames(var = "sample_name")
metadata_wse2 <- metadata_wse %>% column_to_rownames(var = "sample_name")
```

### 2 \| Run PCA w/ `ggpicrust2`

```{r filter, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(ggpicrust2)
library(ggpubr)

## Picrust2::KO PCA - TF only
pathway_pca(
  abundance = KO_tf %>% column_to_rownames("function."), 
  metadata = metadata_tf, 
  group = "season", 
  colors = c("dry" = "lightgreen", "wet"="seagreen4"))

## Picrust2::KO PCA - WS only
pathway_pca(
  abundance = KO_wse %>% column_to_rownames("function."), 
  metadata = metadata_wse, 
  group = "season", 
  colors = c("dry" = "bisque2", "wet"="bisque4"))
```

### 3 \| Run Gene Set Enrichment Analysis w/ `ggpicrust2`

```{r import, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(ggpicrust2)
library(enrichplot)
library(fgsea)

## Prepare GSEA files: pathway type KEGG 
gsea_kegg.tf <- pathway_gsea(
  abundance = KO_tf %>% column_to_rownames("function."),
  metadata = metadata_tf2, 
  group = "season", 
  pathway_type = "KEGG")

gsea_kegg.wse <- pathway_gsea(
  abundance = KO_wse %>% column_to_rownames("function."),
  metadata = metadata_wse2, 
  group = "season", 
  pathway_type = "KEGG")

## Prepare GSEA files: pathway type Gene orthologues
gsea_go.tf <- pathway_gsea(
  abundance = KO_tf %>% column_to_rownames("function."), 
  metadata = metadata_tf2,
  group = "season",  
  pathway_type = "GO", 
  go_category = "BP")

gsea_go.wse <- pathway_gsea(
  abundance = KO_wse %>% column_to_rownames("function."), 
  metadata = metadata_wse2,
  group = "season",  
  pathway_type = "GO", 
  go_category = "BP")

## Annotate GSEA files 
kegg_annotated.tf <- gsea_pathway_annotation(gsea_kegg.tf, pathway_type = "KEGG")
kegg_annotated.wse <- gsea_pathway_annotation(gsea_kegg.wse, pathway_type = "KEGG")

go_annotated.tf <- gsea_pathway_annotation(gsea_go.tf, pathway_type = "GO")
go_annotated.wse <- gsea_pathway_annotation(gsea_go.wse, pathway_type = "GO")

## Drop human microbiome/diseases pathways 
kegg_annotated.tf = kegg_annotated.tf[-(which(kegg_annotated.tf$pathway_name.y %in% c("Alzheimer disease", "Huntington disease","Parkinson disease","Shigellosis", "Bacterial invasion of epithelial cells","Staphylococcus aureus infection"))),]

kegg_annotated.wse = kegg_annotated.wse[-(which(kegg_annotated.wse$pathway_name.y %in% c("Alzheimer disease", "Huntington disease","Parkinson disease","Shigellosis", "Bacterial invasion of epithelial cells","Staphylococcus aureus infection"))),]

## Apply p-value filter (only keep pws that are signif)
kegg_annotated.tf.signif <- kegg_annotated.tf %>% filter(p.adjust < 0.1)
kegg_annotated.wse.signif <- kegg_annotated.wse %>% filter(p.adjust < 0.1)

## Plot GSEA 
visualize_gsea(gsea_results = kegg_annotated.tf.signif, 
               plot_type = "dotplot", 
               n_pathways = 40,
               sort_by = "pvalue", 
               pathway_label_column = "pathway_name.y")

visualize_gsea(gsea_results = kegg_annotated.wse.signif, 
               plot_type = "dotplot", 
               n_pathways = 40,
               sort_by = "pvalue", 
               pathway_label_column = "pathway_name.y")
```
