---
title: "Fig. 3: Relative abundances of putative soil organic matter (SOM)-cycling taxa"
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script is used to analyse the cumulative and individual abundances of putative soil organic matter (SOM)-cycling taxa at the phylum and class level for both bacteria and fungi. Taxa selection was made based on current literature (as of Nov. 2025).

### 1 \| Pre-processing: Define `phyloseq` objects for bacteria and fungi

Prior to analysis, we have to create a phyloseq object from our data. This will be used for most of the downstream analysis.

**Important!** The file 'rooted_tree.nwk' is generated via QIIME2 pre-processing workflow. The detailed script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.**

```{r phyloseq, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load dependencies
library(readxl) # import xlsx files
library(ape)    # import nwk files
library(dplyr)  # data wrangling

## Import data 
otu_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "asv") %>% as.data.frame()
tax_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "tax") %>% as.data.frame()
data_bac <- read_excel("BAC_silva138.2.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_bac <- ape::read.tree("BAC_rooted_tree_dada2.nwk")

otu_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "asv") %>% as.data.frame()
tax_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "tax") %>% as.data.frame()
data_fun <- read_excel("FUN_UNITE10.xlsx", sheet = "meta_all") %>% as.data.frame()
tree_fun <- ape::read.tree("FUN_rooted_tree_dada2.nwk")

## Format variable classes
library(dplyr)  # data wrangling

data_bac$site <- as.factor(data_bac$site)
data_bac$subsite <- as.factor(data_bac$subsite)
data_bac$season <- as.factor(data_bac$season)

data_fun$site <- as.factor(data_fun$site)
data_fun$subsite <- as.factor(data_fun$subsite)
data_fun$season <- as.factor(data_fun$season)

## Adjust taxa names  
tax_bac$Kingdom = gsub("d__","",as.character(tax_bac$Kingdom))
tax_bac$Phylum = gsub("p__","",as.character(tax_bac$Phylum))
tax_bac$Class = gsub("c__","",as.character(tax_bac$Class))
tax_bac$Order = gsub("o__","",as.character(tax_bac$Order))
tax_bac$Family = gsub("f__","",as.character(tax_bac$Family))
tax_bac$Genus = gsub("g__","",as.character(tax_bac$Genus))
tax_bac$Species = gsub("s__","",as.character(tax_bac$Species))
otu_bac <- otu_bac %>% tibble::column_to_rownames("asv")
tax_bac <- tax_bac %>% tibble::column_to_rownames("asv")
data_bac <- data_bac %>% tibble::column_to_rownames("smp_id")

tax_fun$Kingdom = gsub("k__","",as.character(tax_fun$Kingdom))
tax_fun$Phylum = gsub("p__","",as.character(tax_fun$Phylum))
tax_fun$Class = gsub("c__","",as.character(tax_fun$Class))
tax_fun$Order = gsub("o__","",as.character(tax_fun$Order))
tax_fun$Family = gsub("f__","",as.character(tax_fun$Family))
tax_fun$Genus = gsub("g__","",as.character(tax_fun$Genus))
tax_fun$Species = gsub("s__","",as.character(tax_fun$Species))

otu_fun <- otu_fun %>% tibble::column_to_rownames("asv")
tax_fun <- tax_fun %>% tibble::column_to_rownames("asv")
data_fun <- data_fun %>% tibble::column_to_rownames("smp_id")

## Assemble ps objects 
library(phyloseq)

otu_bac <- as.matrix(otu_bac)
tax_bac <- as.matrix(tax_bac)
OTU_B = phyloseq::otu_table(otu_bac, taxa_are_rows = T)
TAX_B = phyloseq::tax_table(tax_bac)
samples_B = phyloseq::sample_data(data_bac)
tree_B = phyloseq::phy_tree(tree_bac)
ps.bac <- phyloseq::phyloseq(OTU_B, TAX_B, samples_B, tree_B)

otu_fun <- as.matrix(otu_fun)
tax_fun <- as.matrix(tax_fun)
OTU_F = phyloseq::otu_table(otu_fun, taxa_are_rows = T)
TAX_F = phyloseq::tax_table(tax_fun)
samples_F = phyloseq::sample_data(data_fun)
tree_F = phyloseq::phy_tree(tree_fun)
ps.fun <- phyloseq::phyloseq(OTU_F, TAX_F, samples_F, tree_F)

ps.bac
ps.fun

## Tidy ps objects w/ microeco
library(microeco)
library(file2meco)

meco.fun = phyloseq2meco(ps.fun)
meco.bac = phyloseq2meco(ps.bac)

meco.bac$tax_table %>% base::subset(Kingdom == "k__Archaea" | Kingdom == "k__Bacteria")
meco.fun$tax_table %>% base::subset(Kingdom == "k__Fungi")

meco.bac$filter_pollution(taxa = c("mitochondria", "chloroplast"))
meco.fun$filter_pollution(taxa = c("mitochondria", "chloroplast"))

meco.bac$tidy_dataset()                                                         
meco.fun$tidy_dataset()

## Convert back to ps
ps.meco.bac = meco2phyloseq(meco.bac)
ps.meco.fun = meco2phyloseq(meco.fun)

ps.bac
ps.meco.bac
ps.fun
ps.meco.fun

## Normalize
library(microbiomeMarker) # TSS normalization
library(dplyr)            # data wrangling

ps.bac.tss = ps.meco.bac %>% microbiomeMarker::normalize(method = "TSS")
ps.fun.tss = ps.meco.fun %>% microbiomeMarker::normalize(method = "TSS")
```

### 2 \| Define putative SOM-cycling taxa

```{r filter, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Bacteria: phylum level
chosen_bac <- c(
  "Pseudomonadota",
  "Bacteroidota",
  "Bacillota",
  "Acidobacteriota",
  "Actinomycetota",
  "Verrucomicrobiota",
  "Chloroflexota",
  "Planctomycetota",
  "Armatimonadota"
)

## Bacteria: class level
chosen_bac2 <- c(
  "Actinobacteria",
  "Alphaproteobacteria",
  "Bacilli",
  "Acidobacteriae",
  "Subgroup_5",
  "Gemmatimonadia",
  "Vicinamibacteria",
  "Chloroflexia",
  "Verrucomicrobiia",
  "Thermoleophilia",
  "Planctomycetes",
  "Acidimicrobiia",
  "Omnitrophia",
  "Phycisphaerae",
  "Chthonomonadia",
  "Armatimonadia",
  "Bacteroidia",
  "Clostridia",
  "Negativicutes",
  "Desulfitobacteriia",
  "Sulfobacillia",
  "Blastocatellia",
  "Holophagae",
  "Thermoanaerobaculia",
  "Ktedonobacteria",
  "TK10",
  "AD3",
  "JG30-KF-CM66",
  "SHA-26",
  "Anaerolineae",
  "Chloroflexia",
  "BD7-11",
  "Pla4_lineage",
  "vadinHA49"
  )

## Fungi: class level
chosen_fun <- c(
  "Agaricomycetes",
  "Sordariomycetes",
  "Eurotiomycetes",
  "Dothideomycetes",
  "Leotiomycetes",
  "Umbelopsidomycetes",
  "Mucoromycetes",
  "Rhizophydiomycetes",
  "Chytridiomycetes",
  "Mortierellomycetes",
  "Spizellomycetes",
  "Neocallimastigomycetes",
  "Monoblepharidomycetes")

## Fungi: order level
chosen_fun2 <- c(
  "Agaricales",
  "Atheliales",
  "Boletales",
  "Hymenochaetales",
  "Polyporales",
  "Russulales",
  "Trechisporales",
  "Thelephorales",
  "Sordariales",
  "Xylariales",
  "Hypocreales",
  "Coniochaetales",
  "Chaetosphaeriales",
  "Amphisphaeriales",
  "Boliniales",
  "Calosphaeriales",
  "Microascales",
  "Diaporthales",
  "Glomerellales",
  "Pleosporales",
  "Capnodiales",
  "Mycosphaerellales",
  "Venturiales",
  "Dothideales",
  "Tubeufiales",
  "Eurotiales",
  "Helotiales",
  "Leotiales",
  "Rhytismatales",
  "Mucorales",
  "Mortierellales",
  "Umbelopsidales"
)
```

### 3 \| Plot cumulative abundance of bacterial and fungal SOM-cycling taxa

#### 3.1 \| Filter microbial data for putative SOM-cyclers

Using the initially created `phyloseq` object, we apply a filter based on the selection of putative SOM-cycling taxa from above, so that only putative taxa that occur in our samples remain. In this particular case, for our data, we found an overabundance of *Gammaproteobacteria* that overshadowed the overarching pattern for other putative SOM-cycling taxa. Hence, we excluded *Gammaproteobacteria* class specifically at a later step to illustrate cumulative abundances of other SOM-cycling taxa to show overall trends.

```{r import, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Extract & prepare the data (Bacteria)
library(phyloseq)
df_bac <- psmelt(ps.bac.tss) %>% dplyr::filter(Sample != "DSTB") 
df_bac$Kingdom <- gsub("^[kpcofgs]__*", "", df_bac$Kingdom)
df_bac$Phylum <- gsub("^[kpcofgs]__*", "", df_bac$Phylum)
df_bac$Class <- gsub("^[kpcofgs]__*", "", df_bac$Class)
df_bac$Order <- gsub("^[kpcofgs]__*", "", df_bac$Order)
df_bac$Family <- gsub("^[kpcofgs]__*", "", df_bac$Family)
df_bac$Genus <- gsub("^[kpcofgs]__*", "", df_bac$Genus)
df_bac$Species <- gsub("^[kpcofgs]__*", "", df_bac$Species)

## Extract & prepare the data (Fungi)
df_fun <- psmelt(ps.fun.tss)
df_fun$Kingdom <- gsub("^[kpcofgs]__*", "", df_fun$Kingdom)
df_fun$Phylum <- gsub("^[kpcofgs]__*", "", df_fun$Phylum)
df_fun$Class <- gsub("^[kpcofgs]__*", "", df_fun$Class)
df_fun$Order <- gsub("^[kpcofgs]__*", "", df_fun$Order)
df_fun$Family <- gsub("^[kpcofgs]__*", "", df_fun$Family)
df_fun$Genus <- gsub("^[kpcofgs]__*", "", df_fun$Genus)
df_fun$Species <- gsub("^[kpcofgs]__*", "", df_fun$Species)

## Extract unique classes (bacteria)
chosen_bac3 = df_bac %>% filter(Phylum %in% chosen_bac, !is.na(Class), Class != "") %>%
  filter(Class != "Gammaproteobacteria") %>%
  distinct(Class) %>%
  pull(Class)
```

#### 3.2 \| Prepare joint abundances df

After importing our data, we'll apply the filter to analyse only putative SOM-cycling taxa, creating a joint bacteria-fungi df for subsequent plotting with `ggplot2`.

```{r prepare, message=FALSE, warning=FALSE, results='hide'}
## Make a new grouping column for OM cyclers
df_bac2 <- df_bac %>% 
  mutate(Group = if_else(Class %in% chosen_bac3, "SOM-cycling taxa", "Other"))

df_fun2 <- df_fun %>%
  mutate(Group = if_else(Order %in% chosen_fun2, "SOM-cycling taxa", "Other"))

## Summarize relative abundance by sample and group
df_bac_summary <- df_bac2 %>% group_by(Sample, site, Group) %>%
  summarise(Abundance = sum(Abundance)) %>% ungroup()

df_fun_summary <- df_fun2 %>% group_by(Sample, site, Group) %>%
  summarise(Abundance = sum(Abundance)) %>% ungroup()

## Add grouping column (bacteria/fungi)
df_bac_summary <- df_bac_summary %>% mutate(Community = "Bacteria")
df_fun_summary    <- df_fun_summary %>% mutate(Community = "Fungi")

## Bind community dfs
df_bac.fun_summary <- bind_rows(df_bac_summary, df_fun_summary)

## Group OM cyclers of each sample for combined community 
df_combined_summary <- df_bac.fun_summary %>% group_by(Sample, Group, site) %>%
  summarise(Total_Abundance = sum(Abundance), .groups = "drop")

## Recalculate relative abundances on base of community sum = 2 (instead of 1)
df_combined_RA <- df_combined_summary %>%
  group_by(Sample) %>%
  mutate(Relative_Abundance = Total_Abundance / sum(Total_Abundance)) %>%
  ungroup()

## Actually plot cumulative abundances
library(ggplot2)
ggplot(df_fun_summary, aes(x = Sample, y = Abundance, fill = Group)) +
  geom_bar(stat = "identity") +
  labs(x = NULL, y = "Relative abundance", fill = "Taxa group") +
  ggpubr::theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = function(x) sprintf("%.0f%%", x * 100)) +
  scale_fill_manual(values=c("SOM-cycling taxa"="bisque2", "Other"="lightgrey")) +
  facet_grid(~site, scales = "free_x")
```

### 4 \| Plot individual abundances of bacterial and fungal SOM-cycling taxa

First, we load dependencies and define a common theme for the plot.

```{r prep2, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## Load dependencies
library(dplyr)
library(ggplot2)
library(forcats)

## Define common theme
## define theme 
theme_bubble <- function(){ 
  theme(legend.key=element_blank(), 
        axis.text.x = element_text(colour = "black", size = 9, angle = 0, vjust = 0.3, hjust = 1), 
        axis.text.y = element_text(colour = "black", face = "italic", size = 9), 
        legend.text = element_text(size = 10, colour ="black"), 
        legend.title = element_text(size = 11, face = "plain"), 
        panel.background = element_blank(), 
        panel.border = element_rect(colour = "black"),
        strip.background = element_blank(),
        legend.position = "right")}
```

Next, we extract the relative abundance means for each taxa at the chosen taxonomic level to plot later.

```{r prep3, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## define parameters
bac_level <- "Class"
fun_level <- "Class"
sample_order <- NULL

## extract means for each taxa at rank
class_mean.fun <- df_fun %>%
  filter(!is.na(Class)) %>%
  group_by(subsite, season, Sample, Class) %>%  
  summarise(sample_RA = sum(Abundance), .groups = "drop") %>%
  group_by(subsite, season, Class) %>%
  summarise(mean_RA = mean(sample_RA) * 100, sd_RA   = sd(sample_RA) * 100, .groups = "drop") %>%
  arrange(subsite, season, desc(mean_RA))

phyla_mean.bac <- df_bac %>%
  mutate(taxa_plot = ifelse(Phylum == "Pseudomonadota" & Class == "Gammaproteobacteria", "Other", Phylum)) %>% 
  filter(!is.na(taxa_plot)) %>%
  group_by(subsite, season, Sample, taxa_plot) %>%   
  summarise(sample_RA = sum(Abundance), .groups = "drop") %>%
  group_by(subsite, season, taxa_plot) %>%
  summarise(mean_RA = mean(sample_RA) * 100, sd_RA   = sd(sample_RA) * 100, .groups = "drop") %>%
  arrange(subsite, season, desc(mean_RA))
```

Next, we bind our data for joint plotting to achieve a joint plot legend, and then plot.

```{r plot, message=FALSE, warning=FALSE, results='hide', tidy=TRUE}
## Steps
## - first we mutate to later bind
## - then we filter taxa
## - then we rename the taxa column to match for successful binding 

## -> Filters
chosen_bac  # phyla filter 
chosen_bac2 # class filter
chosen_fun  # class filter 

## Add shared grouping column
class_mean.fun <- class_mean.fun %>% mutate(Community = "Fungi") %>% 
  filter(Class %in% chosen_fun) %>% rename(taxa = Class)

phyla_mean.bac <- phyla_mean.bac %>% mutate(Community = "Bacteria") %>%
  filter(taxa_plot %in% chosen_bac) %>% rename(taxa = taxa_plot)

## Bind dfs into joint object
taxa_mean.bac.fun <- bind_rows(class_mean.fun, phyla_mean.bac)

## Plot joint df
library(ggplot2)

taxa_mean.bac.fun %>%
  mutate(subsite = recode(subsite, "campina"="C2","campinarana"="C1","plateau"="PL","terrace"="TE")) %>%
  arrange(Community, taxa) %>%              
  mutate(taxa = factor(taxa, levels = unique(taxa))) %>%
  ggplot(aes(x = taxa, y = season, size = mean_RA, fill = season)) +
  geom_point(shape = 21, stroke = 0.6) +
  coord_flip() +
  facet_grid(~subsite)+
  scale_size_continuous(range = c(0.5, 10), labels = function(x) paste0(x, "%")) +
  theme_bubble() +
  labs(y=NULL,x=NULL, size = "mean %RA")
```
