---
title: "Fig. 5: Linking taxonomic structure and predicted functional potential"
author: "Jessica Finck"
date: "2026-02-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

This script is used to analyse the composition-function coupling for bacterial communities, using `PICRUSt2` output, feature table and taxonomy data.

**Important!** As specified in the manuscript, we removed sample `DSTB` due to likely contamination. The exclusion step is, henceforth, fully optional when replicating this script.

### 1 \| Pre-processing: Load necessary taxonomic and functional data

#### 1.1 \| Load data for Mantel test

```{r data1, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Import data 
library(readxl)
library(tibble)
library(dplyr)
bac_asv <- read_xlsx("BAC_silva138.2.xlsx", sheet = "asv_inv") %>% 
  as.data.frame() %>% filter(asv!="DSTB")

bac_dat <- read_xlsx("BAC_silva138.2.xlsx", sheet = "meta_all") %>% 
  as.data.frame() %>% filter(smp_id != "DSTB")

bac_KO = read.delim("KO_per_sample.tsv", row.names = 1, check.names = FALSE) %>% select(-"DSTB")

## Format datalibrary(janitor)
library(janitor)
bac_asv <- bac_asv %>% tibble::column_to_rownames("asv") 
bac_KO = bac_KO %>% t() 

## Subset data
tf_samples  <- bac_dat$smp_id[bac_dat$site == "TF"] %>% as.list()
bac_KO_tf <- bac_KO[row.names(bac_KO) %in% tf_samples,] 
bac_asv_tf <- bac_asv[row.names(bac_asv) %in% tf_samples,]
bac_dat_tf <- bac_dat[bac_dat$smp_id %in% tf_samples,]

wse_samples <- bac_dat$smp_id[bac_dat$site == "WSE"]
bac_KO_wse <- bac_KO[row.names(bac_KO) %in% wse_samples,]
bac_asv_wse <- bac_asv[row.names(bac_asv) %in% wse_samples,] 
bac_dat_wse <- bac_dat[bac_dat$smp_id %in% wse_samples,]

## Ensure samples match and are in same order:
common_asv.ko_tf <- intersect(rownames(bac_asv_tf), rownames(bac_KO_tf))
common_asv.ko_wse <- intersect(rownames(bac_asv_wse), rownames(bac_KO_wse))

## Reorder rows to be the same order
## ... terra firme 
bac_asv_tf  <- bac_asv_tf[common_asv.ko_tf, , drop = FALSE]
bac_KO_tf <- bac_KO_tf[common_asv.ko_tf, , drop = FALSE]

bac_dat.ko_tf <- bac_dat %>% 
  filter(smp_id %in% common_asv.ko_tf) %>% 
  column_to_rownames("smp_id")
# reorder to match matrices
bac_dat.ko_tf <- bac_dat.ko_tf[common_asv.ko_tf, , drop = FALSE]

## ... white-sand
bac_asv_wse  <- bac_asv_wse[common_asv.ko_wse, , drop = FALSE]
bac_KO_wse <- bac_KO_wse[common_asv.ko_wse, , drop = FALSE]
bac_dat.ko_wse <- bac_dat %>% 
  filter(smp_id %in% common_asv.ko_wse) %>% 
  column_to_rownames("smp_id")
# reorder to match matrices
bac_dat.ko_wse <- bac_dat.ko_wse[common_asv.ko_wse, , drop = FALSE]  
```

After preparing the data, we calculate the necessary distance matrix. Here, we chose *Bray-Curtis dissimilarity* matrix, which is calculated based on normalized (i.e. total sum scaled) read counts. Another way to do it is via *robust Aitchison* (rAitchison) distance, which requires raw counts to be transformed with *robust CLR transformation*. In either case, we are using the `decostand()` function of the `vegan` package. The key difference is that Bray-Curtis distances are non-euclidean, so the resulting ordination is a PCoA, while for rAitchison the transformation brings the data into an euclidean space, rendering the resulting ordinations euclidean. Both are viable options.

```{r distances, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(vegan)

## Calculate rAitchison dist
dist_asv_tf.rAit  <- bac_asv_tf %>% 
  decostand(method = "rclr") %>% 
  vegdist(method = "euclidean")  
dist_ko_tf.rAit <- bac_KO_tf %>% 
  decostand(method = "rclr") %>% 
  vegdist(method = "euclidean")

dist_asv_wse.rAit <- bac_asv_wse %>% 
  decostand(method = "rclr") %>% 
  vegdist(method = "euclidean")
dist_ko_wse.rAit <- bac_KO_wse %>% 
  decostand(method = "rclr") %>% 
  vegdist(method = "euclidean")

## Calculate Bray-Curtis dist
dist_asv_tf.bray  <- bac_asv_tf %>% 
  decostand(method = "total", MARGIN = 1) %>% 
  vegdist(method = "bray")
dist_ko_tf.bray <- bac_KO_tf %>% 
  decostand(method = "total", MARGIN = 2) %>% 
  vegdist(method = "bray")

dist_asv_wse.bray  <- bac_asv_wse %>% 
  decostand(method = "total", MARGIN = 1) %>% 
  vegdist(method = "bray")
dist_ko_wse.bray <- bac_KO_wse %>% 
  decostand(method = "total", MARGIN = 2) %>% 
  vegdist(method = "bray")

## Sanity check: sample labels match between distance objects
stopifnot(all(labels(dist_asv_tf.bray) == labels(dist_ko_tf.bray)))
stopifnot(all(labels(dist_asv_wse.bray) == labels(dist_ko_wse.bray)))

stopifnot(all(labels(dist_asv_tf.rAit) == labels(dist_ko_tf.rAit)))
stopifnot(all(labels(dist_asv_wse.rAit) == labels(dist_ko_wse.rAit)))
```

#### 1.2 \| Load data for Procrustes analysis

Hypothetically, this is the exact same procedure are for loading and preparing the Mantel test objects, with the exception that we – specifically – excluded sample `DSTB` for Mantel tests but kept it for Procrustes analysis to show a fuller picture. Hence, we reload the "Load data for Mantel test" but modified object names to keep both within the R markdown environment.

```{r data2, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Import data 
library(readxl)
library(tibble)
library(dplyr)
bac_asv2 <- read_xlsx("BAC_silva138.2.xlsx", sheet = "asv_inv") %>% 
  as.data.frame()

bac_dat2 <- read_xlsx("BAC_silva138.2.xlsx", sheet = "meta_all") %>% 
  as.data.frame() 

bac_KO2 = read.delim("KO_per_sample.tsv", row.names = 1, check.names = FALSE)

## Format datalibrary(janitor)
library(janitor)
bac_asv2 <- bac_asv2 %>% tibble::column_to_rownames("asv") 
bac_KO2 = bac_KO2 %>% t() 

## Subset data
tf_samples2  <- bac_dat2$smp_id[bac_dat2$site == "TF"] %>% as.list()
bac_KO_tf2 <- bac_KO2[row.names(bac_KO2) %in% tf_samples2,] 
bac_asv_tf2 <- bac_asv2[row.names(bac_asv2) %in% tf_samples2,]
bac_dat_tf2 <- bac_dat2[bac_dat2$smp_id %in% tf_samples2,]

wse_samples2 <- bac_dat2$smp_id[bac_dat2$site == "WSE"]
bac_KO_wse2 <- bac_KO2[row.names(bac_KO2) %in% wse_samples2,]
bac_asv_wse2 <- bac_asv2[row.names(bac_asv2) %in% wse_samples2,] 
bac_dat_wse2 <- bac_dat2[bac_dat2$smp_id %in% wse_samples2,]

## Ensure samples match and are in same order:
common_asv.ko_tf2 <- intersect(rownames(bac_asv_tf2), rownames(bac_KO_tf2))
common_asv.ko_wse2 <- intersect(rownames(bac_asv_wse2), rownames(bac_KO_wse2))

## Reorder rows to be the same order
## ... terra firme 
bac_asv_tf2  <- bac_asv_tf2[common_asv.ko_tf2, , drop = FALSE]
bac_KO_tf2 <- bac_KO_tf2[common_asv.ko_tf2, , drop = FALSE]

bac_dat.ko_tf2 <- bac_dat2 %>% 
  filter(smp_id %in% common_asv.ko_tf2) %>% 
  column_to_rownames("smp_id")
# reorder to match matrices
bac_dat.ko_tf2 <- bac_dat.ko_tf2[common_asv.ko_tf2, , drop = FALSE]

## ... white-sand
bac_asv_wse2  <- bac_asv_wse2[common_asv.ko_wse2, , drop = FALSE]
bac_KO_wse2 <- bac_KO_wse2[common_asv.ko_wse2, , drop = FALSE]
bac_dat.ko_wse2 <- bac_dat2 %>% 
  filter(smp_id %in% common_asv.ko_wse2) %>% 
  column_to_rownames("smp_id")
# reorder to match matrices
bac_dat.ko_wse2 <- bac_dat.ko_wse2[common_asv.ko_wse2, , drop = FALSE]  

## Calculate Bray-Curtis dist
library(vegan)
dist_asv_tf.bray2  <- bac_asv_tf2 %>% 
  decostand(method = "total", MARGIN = 1) %>% 
  vegdist(method = "bray")
dist_ko_tf.bray2 <- bac_KO_tf2 %>% 
  decostand(method = "total", MARGIN = 2) %>% 
  vegdist(method = "bray")

dist_asv_wse.bray2  <- bac_asv_wse2 %>% 
  decostand(method = "total", MARGIN = 1) %>% 
  vegdist(method = "bray")
dist_ko_wse.bray2 <- bac_KO_wse2 %>% 
  decostand(method = "total", MARGIN = 2) %>% 
  vegdist(method = "bray")
```

#### 1.3 \| Load data for cumulative nutrient cycling vs. drought response potential

**Important!** PICRUSt2 files were generated within the QIIME2 environment using the `picrust2.py` pipeline. The script can be [**found here**](https://github.com/JFinck1994/general_scripts/blob/main/qiime2_amplicon-16S-PE-read-processing.sh)**.**

```{r data3, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(readxl)
library(dplyr)
bac_dat <- read_xlsx("BAC_silva138.2.xlsx", sheet = "meta_all") %>% 
  as.data.frame() %>% 
  filter(smp_id != "DSTB")
bac_KO = read.delim("KO_per_sample.tsv", row.names = 1, check.names = FALSE) %>%
  dplyr::select(-"DSTB")

## Normalize data
## ! Important: Rows must be samples
library(janitor)
library(vegan)
bac_KO = as.data.frame(x = t(bac_KO), stringsAsFactors = FALSE)
bac_KO = decostand(bac_KO, method = "total")
## check if normalized correctly (rows == 1, columns =! 1)
rowSums(bac_KO)
colSums(bac_KO)

## Define subset data based on forest type
tf_samples  <- bac_dat$smp_id[bac_dat$site == "TF"] %>% as.list()
ws_samples <- bac_dat$smp_id[bac_dat$site == "WSE"]

## Subset data 
bac_KO_tf <- bac_KO[row.names(bac_KO) %in% tf_samples,] 
bac_KO_ws <- bac_KO[row.names(bac_KO) %in% ws_samples,]

bac_dat_tf <- bac_dat[bac_dat$smp_id %in% tf_samples,]
bac_dat_ws <- bac_dat[bac_dat$smp_id %in% ws_samples,]
```

### 2 \| Mantel test between taxonomic and functional composition

```{r filter, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
set.seed(123)

## Run Mantel test (rAitch)
## - pearson and spearman are both viable options since we work with euclidean dist
mantel(dist_asv_tf.rAit, 
       dist_ko_tf.rAit,
       method = "spearman", 
       permutations = 999)

mantel(dist_asv_wse.rAit, 
       dist_ko_wse.rAit, 
       method = "spearman", 
       permutations = 999)

## Run Mantel test (Bray-Curtis)
## - spearman only since these are non-euclidean dist
mantel(dist_asv_tf.bray, 
       dist_ko_tf.bray,
       method = "spearman", 
       permutations = 999)

mantel(dist_asv_wse.bray, 
       dist_ko_wse.bray, 
       method = "spearman", 
       permutations = 999)

## Define Mantel results as object for plotting 
mantel_tf.bray <- data.frame(dist_asv_tf.bray = as.vector(dist_asv_tf.bray),
                             dist_ko_tf.bray = as.vector(dist_ko_tf.bray))
mantel_ws.bray <- data.frame(dist_asv_wse.bray = as.vector(dist_asv_wse.bray),
                             dist_ko_wse.bray = as.vector(dist_ko_wse.bray))

## Visualize Mantel results 
library(ggplot2)
library(scales)

ggplot(mantel_tf.bray, aes(x = dist_asv_tf.bray, y = dist_ko_tf.bray)) + 
  geom_point(size = 3, alpha = 0.4) + 
  geom_smooth(method = "lm", color = "blue") + 
  ggpubr::theme_pubr() +
  labs(x = "Taxonomic dissimilarity", y = "Functional dissimilarity") +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 0.1, decimal.mark = '.'), 
    breaks = c(0,0.2,0.4,0.6,0.8)) +
  scale_x_continuous(
    labels = scales::number_format(accuracy = 0.1, decimal.mark = '.'), 
    breaks = c(0.2,0.4,0.6,0.8,1.0)) +
  annotate("text", x = 0.5, y = -0.15, 
           label = paste0("Mantel r = ", round(0.7443, 3),", p < 0.01"), 
           hjust = 0.8, vjust = 0.2, size = 4, fontface = "italic") +
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        panel.border = element_rect(colour = "black", fill=NA))

ggplot(mantel_ws.bray, aes(x = dist_asv_wse.bray, y = dist_ko_wse.bray)) +
  geom_point(size = 3, alpha = 0.4) + 
  geom_smooth(method = "lm", color = "blue") + 
  ggpubr::theme_pubr() +
  labs(x = "Taxonomic dissimilarity", y = "Functional dissimilarity") +
  scale_y_continuous(
    labels = scales::number_format(accuracy = 0.1, decimal.mark = '.'), 
    breaks = c(0,0.2,0.4,0.6,0.8)) +
  scale_x_continuous(
    labels = scales::number_format(accuracy = 0.1, decimal.mark = '.'), 
    breaks = c(0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9)) +
  annotate("text", x = 0.5, y = -0.1, 
           label = paste0("Mantel r = ", round(0.5907, 3),", p < 0.001"), 
           hjust = 0.65, vjust = 0.2, size = 4, fontface = "italic") +
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        panel.border = element_rect(colour = "black", fill=NA))
```

### 3 \| Procrustes test / ordination concordance

Similar to Mantel tests, Procrustes analysis can be performed w/ PCoA or PCA, depending on the distance matrix. Here, we chose *Bray-Curtis dissimilarity* matrix, which is also what the script is aimed for. Using the PCA/rAitchison distance from (1.1) changes the script a little bit. Instead of using the `pcoa()` function of the `ape` package, we are using the `prcomp()` function to calculate PCA ordinations (check alternative code below). In this case, the `pcoa_*` objects have to be replaced with `pca_*`.

`pca_asv_tf  <- bac_asv_tf %>% decostand(method = "rclr") %>% prcomp()`

`pca_ko_tf <- bac_KO_tf %>% decostand(method = "rclr") %>% prcomp()`

`pca_asv_wse  <- bac_asv_wse %>% decostand(method = "rclr") %>% prcomp()`

`pca_ko_wse <- bac_KO_wse %>% decostand(method = "rclr") %>% prcomp()`

```{r procrustes, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Create PCoA for Procrustes
library(ape)
pcoa_asv_tf  <- pcoa(dist_asv_tf.bray2)
pcoa_ko_tf <- pcoa(dist_ko_tf.bray2)
pcoa_asv_wse  <- pcoa(dist_asv_wse.bray2)
pcoa_ko_wse <- pcoa(dist_ko_wse.bray2)

## Extract first two PCoA axes
coords_asv_tf  <- as.data.frame(pcoa_asv_tf$vectors[, 1:2])
coords_ko_tf <- as.data.frame(pcoa_ko_tf$vectors[, 1:2])
coords_asv_wse  <- as.data.frame(pcoa_asv_wse$vectors[, 1:2])
coords_ko_wse <- as.data.frame(pcoa_ko_wse$vectors[, 1:2])

## Rename columns
colnames(coords_asv_tf)  <- c("PCoA1_asv", "PCoA2_asv")
colnames(coords_ko_tf) <- c("PCoA1_ko", "PCoA2_ko")
colnames(coords_asv_wse)  <- c("PCoA1_asv", "PCoA2_asv")
colnames(coords_ko_wse) <- c("PCoA1_ko", "PCoA2_ko")

coords_asv_tf$smp_id  <- rownames(bac_asv_tf2)
coords_ko_tf$smp_id <- rownames(bac_KO_tf2)
coords_asv_wse$smp_id  <- rownames(bac_asv_wse2)
coords_ko_wse$smp_id <- rownames(bac_KO_wse2)

## Merge with metadata
coords_asv.ko_tf <- left_join(coords_asv_tf, coords_ko_tf, by = "smp_id")
coords_asv.ko_tf <- left_join(coords_asv.ko_tf, bac_dat2, by = "smp_id")
coords_asv.ko_wse <- left_join(coords_asv_wse, coords_ko_wse, by = "smp_id")
coords_asv.ko_wse <- left_join(coords_asv.ko_wse, bac_dat2, by = "smp_id")
 
## prepare PCoA matrices 
A_tf <- as.matrix(coords_asv_tf[, c("PCoA1_asv", "PCoA2_asv")])
B_tf <- as.matrix(coords_ko_tf[, c("PCoA1_ko", "PCoA2_ko")])
rownames(A_tf) <- coords_asv_tf$smp_id
rownames(B_tf) <- coords_ko_tf$smp_id
procrustes_tf <- procrustes(A_tf, B_tf, scale = TRUE)
set.seed(123)
protest_tf <- protest(A_tf, B_tf, permutations = 999)

A_wse <- as.matrix(coords_asv_wse[, c("PCoA1_asv", "PCoA2_asv")])
B_wse <- as.matrix(coords_ko_wse[, c("PCoA1_ko", "PCoA2_ko")])
rownames(A_wse) <- coords_asv_wse$smp_id
rownames(B_wse) <- coords_ko_wse$smp_id
procrustes_wse <- procrustes(A_wse, B_wse, scale = TRUE)
set.seed(123)
protest_wse <- protest(A_wse, B_wse, permutations = 999)

## summary protest / procrustes
summary(protest_tf)
summary(procrustes_tf)
summary(protest_wse)
summary(procrustes_wse)

## prepare for plotting 
fittedB_tf <- as.data.frame(procrustes_tf$Yrot)
colnames(fittedB_tf) <- c("PCoA1_func_rot", "PCoA2_func_rot")
fittedB_tf$smp_id <- rownames(fittedB_tf)

fittedB_wse <- as.data.frame(procrustes_wse$Yrot)
colnames(fittedB_wse) <- c("PCoA1_func_rot", "PCoA2_func_rot")
fittedB_wse$smp_id <- rownames(fittedB_wse)

plotdf_tf <- coords_asv.ko_tf %>% left_join(fittedB_tf, by = "smp_id")
plotdf_wse <- coords_asv.ko_wse %>% left_join(fittedB_wse, by = "smp_id")

## Plot Procrustes:: Terra firme
library(ggplot2)
library(ggrepel)
library(ggpubr)

ggplot(plotdf_tf, aes(x = PCoA1_asv, y = PCoA2_asv)) +
  geom_segment(aes(xend = PCoA1_func_rot, yend = PCoA2_func_rot), 
               color = "black", 
               arrow = arrow(length = unit(0.2, "cm"))) +
  geom_point(aes(fill = season), 
             shape = 24, 
             color = "black", 
             size = 4, 
             stroke = 0.8) +
  geom_point(aes(x = PCoA1_func_rot, y = PCoA2_func_rot, fill = season), 
             shape = 21, 
             color = "black", 
             size = 4, 
             stroke = 0.8) +
  scale_fill_manual(values = c("dry" = "lightgreen", "wet" = "forestgreen")) +
  scale_color_manual(values = c("dry" = "lightgreen", "wet" = "forestgreen")) +
  ggpubr::theme_pubr() + 
  labs(x = "Dimension 1", y = "Dimension 2", fill = "Season") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text_repel(aes(label = smp_id), size = 2.5, max.overlaps = 12) +
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        panel.border = element_rect(colour = "black", fill=NA), 
        legend.position = "none") +
  annotate("text", x = -0.55, y = 0.55, 
           label = paste0("PROTEST \n m² = ", 
                          round(protest_tf$ss, 3),
                          "\np =", signif(protest_tf$signif,3)), 
           hjust = 0.2, vjust = 1, size = 4, fontface = "italic") 

ggplot(plotdf_wse, aes(x = PCoA1_asv, y = PCoA2_asv)) +
  geom_segment(aes(xend = PCoA1_func_rot, yend = PCoA2_func_rot, color = subsite), 
               linewidth = 0.6, alpha = 0.8) +
  geom_point(aes(fill = season), 
             shape = 24, 
             color = "black", 
             size = 4, 
             stroke = 0.9) +
  geom_point(aes(x = PCoA1_func_rot, y = PCoA2_func_rot, fill = season), 
             shape = 21, 
             color = "black", 
             size = 4, 
             stroke = 0.8) +
  scale_fill_manual(values = c("dry" = "bisque1", "wet" = "bisque4")) +
  scale_color_manual(values = c("dry" = "bisque1", "wet" = "bisque4")) +
  ggpubr::theme_pubr() + 
  labs(x = "Dimsension 1", y = "Dimension 2", fill = "Season") +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.1, decimal.mark = '.')) +
  geom_text_repel(aes(label = smp_id), size = 2.5, max.overlaps = 12) +
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        panel.border = element_rect(colour = "black", fill=NA), 
        legend.position = "none") +
  annotate("text", x = -0.4, y = -0.25, 
           label = paste0("PROTEST \n m² = ", 
                          round(protest_wse$ss, 3),
                          "\np =", signif(protest_wse$signif,3)), 
           hjust = 0.2, vjust = 1, size = 4, fontface = "italic") 
```

### 4 \| Cumulative nutrient cycling vs. drought response potential

First, we need to filter our PICRUSt2-derived KO data to only keep KOs of interest (in our case K numbers central to nutrient cycling and drought response and molecule transport pathways acc. to `KO_core_genes`).

```{r NCDR1, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Load K number filter
library(readxl)
KO_core_genes <- read_excel("KO_core_genes.XLSX", sheet = "combined_genes_DOM")

## Create long format data w/ KOs as new column (for later filter)
library(tidyverse)
bac_KO_tf.long <- bac_KO_tf %>%
  rownames_to_column("smp_id") %>% 
  pivot_longer(cols = -smp_id, names_to = "k_number", values_to = "rel_abund")

bac_KO_ws.long <- bac_KO_ws %>%
  rownames_to_column("smp_id") %>%
  pivot_longer(cols = -smp_id, names_to = "k_number", values_to = "rel_abund")

## Apply filter
bac_KO_tf.filt <- bac_KO_tf.long %>% semi_join(KO_core_genes, by = "k_number")
bac_KO_ws.filt <- bac_KO_ws.long %>% semi_join(KO_core_genes, by = "k_number")
```

Next, we need to join our functional data w/ our metadata, and functional group assignments, aka which K numbers belong to which group (nutrient cycling or drought response).

**Important!** For joining w/ the functional groups, the selected columns are the categorizations for the different K numbers that can be found in `KO_core_genes`.

```{r NCDR2, tidy=TRUE, message=FALSE, warning=FALSE, results='hide'}
## Join w/ metadata/season
bac_KO_tf.filt <- bac_KO_tf.filt %>% left_join(bac_dat_tf %>% select(smp_id, season), by = "smp_id")
bac_KO_ws.filt <- bac_KO_ws.filt %>% left_join(bac_dat_ws %>% select(smp_id, season), by = "smp_id")

## Join w/ metadata/subsite
bac_KO_tf.filt <- bac_KO_tf.filt %>% left_join(bac_dat_tf %>% select(smp_id, subsite), by = "smp_id")
bac_KO_ws.filt <- bac_KO_ws.filt %>% left_join(bac_dat_ws %>% select(smp_id, subsite), by = "smp_id")

## Join w/ functional groups
bac_KO_tf.filt <- bac_KO_tf.filt %>% 
  left_join(KO_core_genes %>% 
              select(k_number, bar_filter, bar_group, bar_pathway, z_group1, z_group2, z_group3), by = "k_number")

bac_KO_ws.filt <- bac_KO_ws.filt %>% 
  left_join(KO_core_genes %>% 
              select(k_number, bar_filter, bar_group, bar_pathway, z_group1, z_group2, z_group3), by = "k_number")
```

Finally, we prepare our data for plotting by summarizing at the chosen pathway/functional group level, and then binding together the dfs of both forest types for a joint plot.

```{r NCDR3, tidy=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(tidyr)

## Write function to summarize multiple grouping variables
summarize_abundance <- function(
    df, 
    group_vars, 
    value_col = "rel_abund", 
    sample_col = "smp_id", 
    season_col = "season", 
    site_col = "subsite") {
  
  summary_list <- list()
  
  for (g in group_vars) {
    
    temp <- df %>%
      group_by(
        !!sym(sample_col), 
        !!sym(season_col), 
        !!sym(site_col), 
        !!sym(g)) %>%
      summarise(rel_abund = sum(!!sym(value_col)), .groups = "drop") %>%
      rename(group = !!sym(g)) %>%
      mutate(group_type = g)
    
    summary_list[[g]] <- temp
  }
  
  bind_rows(summary_list)
}

## Define the grouping variables you want summaries for
group_vars <- c("z_group1", "z_group2", "z_group3","bar_filter", "bar_pathway", "bar_group")

## Write summary dfs for each forest type 
summary_KO.tf <- summarize_abundance(bac_KO_tf.filt, group_vars)
summary_KO.ws <- summarize_abundance(bac_KO_ws.filt, group_vars)

## Combine into 1 joint summary df w/ new forest_type column
summary_KO.tf <- summary_KO.tf %>% mutate(forest_type = "Terra firme")
summary_KO.ws <- summary_KO.ws %>% mutate(forest_type = "White-sand")
summary_KO_all <- bind_rows(summary_KO.tf, summary_KO.ws)

## Plot as barplot
library(ggpubr)
summary_KO_all %>% 
  filter(group_type == "bar_filter") %>% 
  ggbarplot("group", "rel_abund",
            fill = "season", color = "season", 
            palette = "Paired",
            position = position_dodge(0.8), 
            add = c("mean_sd"), 
            error.plot = "upper_errorbar",
            facet.by = "forest_type") +
  scale_y_continuous(labels = function(x) sprintf("%.1f", x * 100)) +
  labs(y = "Mean %RA", x = NULL) +
  stat_compare_means(aes(group = season), label = "p.signif", hide.ns = TRUE) +
  geom_jitter(aes(y=rel_abund, color = season), size = 3, width = 0.1, alpha = 0.7) +
  theme(strip.background = element_blank(), 
        strip.text = element_text(size = 12, hjust = 0), 
        axis.text.x = element_text(angle = 45, hjust = 1))
```
